---
title: "Portulaca Ancestral Reconstruction"
author: "Nora Heaphy"
date: "11/15/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ape)
library(ggtree)
library(phytools)
library(tidyverse)
library(ggplot2)
library(phylotools)
library(knitr)

```

## Data cleaning

- First to get all the taxa names in the character data matrix and the phylogeny tip labels into the format *Genus_species*

- Then drop the tips from the tree that don't have an equivalent in the character data (which is many of them) with `drop.tip`

- Then remove the character data rows that don't have an equivalent in the tree with `dplyr`

- Then use this line from exercise 10 to get the character data rows into the right order to map onto the tips: `geoData = geoData[ match( row.names(geoData), geoTree$tip.label  ), ]`

```{r}

# read in tree file
portulacaTree <- read.tree("data/Ocampo_Portulaca.parts.treefile")

# read in character data
portulacaData <- read.csv("data/portulaca_means_11-06-2020.csv", header = TRUE, 
                          sep = ",", na.strings = "")

portulacaTree$tip.label = portulacaTree$tip.label %>%
  str_replace("-.*","") %>%
  str_replace("_", " ")

# make dataframe of species names from character data
species_character <- data.frame(portulacaData$species) %>%
  rename(names = portulacaData.species)

# make dataframe of species names from tip labels
species_tree <- data.frame(portulacaTree$tip.label) %>%
  rename(names = portulacaTree.tip.label)

# pull out the names in the tree but not the character data
names_to_drop <- setdiff(species_tree, species_character)
names_to_drop <- names_to_drop$names # make into a vector

# remove tips that don't have character data matches
portulacaTree <- drop.tip(portulacaTree, names_to_drop)
portulacaTree <- drop.tip(portulacaTree, c(9, 4)) # monophyletic duplicates

# now remove rows from character data frame that don't correspond to tips
species_tree <- data.frame(portulacaTree$tip.label) %>%
  rename(names = portulacaTree.tip.label)

names_to_drop <- setdiff(species_character, species_tree)
names_to_drop <- names_to_drop$names # make into a vector

portulacaData <- portulacaData[!(portulacaData$species %in% names_to_drop),] 

# remove P. halimoides and P. pusilla from both tree and character data, because
# they're resolving as non-monophyletic in the Ocampo & Columbus phylogeny
portulacaTree <- drop.tip(portulacaTree, c("Portulaca halimoides", "Portulaca pusilla"))
portulacaData <- portulacaData %>% filter(species != "Portulaca halimoides" &
                                            species != "Portulaca pusilla")

# last check


if(! length(portulacaTree$tip.label) == nrow(portulacaData)) {
      stop("Error: different numbers of taxa in tip labels and character data!")
} else {
  print("All good, continue!")
}

# save this cleaned character data
write.csv(portulacaData, "data/portulaca_means_mapping.csv", row.names = FALSE)

```


```{r fig.width=9, fig.height=18}

# plot tree

plot(portulacaTree)

```

```{r}

# turn first column of character data into row names
row.names(portulacaData) = portulacaData$species

# ensure character data and tiplabels are in same order
portulacaData <- portulacaData[match(portulacaTree$tip.label,row.names(portulacaData)), ]
    
# double check that labels are the same, in case some were missing or duplicated
if(!all(row.names(portulacaData) == portulacaTree$tip.label)){
      stop("Error: tip labels and dataframe rows not in same order!")
} else {
  print("Labels are in the right order, go ahead!")
}

# count the number of tips
n_tips = length(portulacaTree$tip.label)

# root tree manually based on Ocampo & Columbus
portulacaTree <- root(portulacaTree, node = 28, edgelabel = TRUE)
if(is.rooted(portulacaTree) == FALSE) {
  stop("Error: tree is not rooted.")
} else {
  print("Tree is rooted, continue!")
}

# get the root node by finding the node that is a parent (in column 1 of edge) 
# but not a child (in column 2 of edge)
# root_node = portulacaTree$edge[(!portulacaTree$edge[,1] %in% portulacaTree$edge[,2]),1] %>%
#  unique()
root_node = 28 # temporarily assigning manually because I can't get it to work the other way

# get the maximum distance from the root to another node, which will be for a tip.
# since this tree happens to be ultrametric, this will be the root age
root_age = max(dist.nodes(portulacaTree)[, root_node])

# this isn't necessary, but since we don't know the units of edge length let's rescale
# the tree so that the age of the root is 1.
# this could be a place to add noise into edge lengths to see how robust the analysis is
portulacaTree$edge.length = portulacaTree$edge.length/root_age
    
# and now recalculate the root age to make sure things look good
root_age = max(dist.nodes(portulacaTree)[, root_node])
    
```

```{r}

# scale and graph tree

label_offset = 0.02 * root_age
tree_width = 1.5 * root_age

portulaca_ggtree = ggtree(portulacaTree) + 
  geom_tiplab(fontface = "italic", offset = label_offset) +
  xlim(0, tree_width) +
  geom_text2(aes(label = node), col = "red", nudge_x = label_offset/2, size = 2)

portulaca_ggtree

### for final figure

portulaca_ggtree_fig = ggtree(portulacaTree) + 
  geom_tiplab(fontface = "italic", offset = label_offset) +
  xlim(0, tree_width) +
  ggtitle("Portulaca Phylogeny")

portulaca_ggtree_fig

```

## Plot character data onto tips

```{r}

# to display the node values, ggtree expects a vector where each element
# corresponds to each node in the tree, both at the tips and at the internal 
# nodes. Since we don't have values for the internal nodes, we will pad them 
# with NA (Not Available) for now.
node_values = c(portulacaData$MAT, rep(NA, n_tips-1))

# read in color palette
mycol  <- read.csv("data/color_palette.csv", header = TRUE, sep = ",", na.strings = "")

portulaca_ggtree + 
  geom_tippoint(aes(color = node_values), size = 3, alpha = 1) +
  scale_color_gradientn(colors = mycol$x) +
  ggtitle("Portulaca Phylogeny")

```

## Ancestral character state estimation using Reduced Maximum Likelihood

```{r}

# fix the edge length of 0 by setting it to a very tiny number
portulacaTree$edge.length[portulacaTree$edge.length==0] = 1e-6

mat_ancestral = 
  ace(portulacaData$MAT, 
    portulacaTree, 
    type = "continuous", 
    method = "REML", 
    model = "BM"  
  )

mat_ancestral

```


## Plot reconstructed states on phylogeny

```{r}

# To display the node values, ggtree expects a vector where each element
# corresponds to each node in the tree, both at the tips and at the internal 
# nodes. Since we don't have values for the internal nodes, we will pad them 
# with NA (Not Available) for now.

node_values = c(portulacaData[,22], mat_ancestral$ace)

portulaca_ggtree + 
  geom_tippoint(aes(color = node_values), size = 3, alpha = 1 ) +
  geom_nodepoint(aes(color = node_values), size = 3, alpha = 1 ) +
  scale_color_gradientn(colors = mycol$x, limits = c(9, 27))

```

## Ancestral character state estimation using Phylogenetic Independent Contrasts

```{r}

mat_ancestral_pic = 
  ace(portulacaData$MAT, 
    portulacaTree, 
    type="continuous", 
    method="pic", 
    model="BM"  
  )

mat_ancestral_pic

```



## Plot reconstructed states on phylogeny

```{r}

# To display the node values, ggtree expects a vector where each element
# corresponds to each node in the tree, both at the tips and at the internal 
# nodes. Since we don't have values for the internal nodes, we will pad them 
# with NA (Not Available) for now.
node_values = c(portulacaData[,22], mat_ancestral_pic$ace )

portulaca_ggtree + 
  geom_tippoint(aes(color = node_values), size = 3, alpha = 1 ) +
  geom_nodepoint(aes(color = node_values), size = 3, alpha = 1 ) +
  scale_color_gradientn(colors = mycol$x, limits = c(9, 27))

```

## Compare REML and PIC methods

```{r}

D_methods = data.frame(
  reml = mat_ancestral$ace,
  pic = mat_ancestral_pic$ace
)

# Add node numbers to the data frame
D_methods$node = seq(1:nrow(D_methods)) + n_tips

D_methods %>%
  ggplot(aes(x = reml,y = pic)) +
  geom_point() +
  geom_text(aes(label = node), color = "red", vjust = 0, nudge_y = 0.005) +
  geom_abline(slope = 1, intercept = 0, color = "blue")

# Add confidence intervals to dataframe
D_methods$CI_reml_lower <- mat_ancestral$CI95[,1]
D_methods$CI_reml_upper <- mat_ancestral$CI95[,2]
D_methods$CI_pic_lower <- mat_ancestral_pic$CI95[,1]
D_methods$CI_pic_upper <- mat_ancestral_pic$CI95[,2]

# print in table
kable(D_methods)

```


## Confidence and node age

The code below explores the width of the confidence intervals for each reconstruction.

```{r}

# for REML
interval_width = mat_ancestral$CI95[,2] - mat_ancestral$CI95[,1]
node_values = c(rep(NA, nrow(portulacaData)), interval_width)

portulaca_ggtree + 
  geom_nodepoint(aes(color = node_values), size = 3, alpha = 1) +
  scale_color_gradientn(colors = mycol$x)

# for PIC
interval_width = mat_ancestral_pic$CI95[,2] - mat_ancestral_pic$CI95[,1]
node_values = c(rep(NA, nrow(portulacaData)), interval_width)

portulaca_ggtree + 
  geom_nodepoint(aes(color = node_values), size = 3, alpha = 1) +
  scale_color_gradientn(colors = mycol$x)

# for the difference between the two reconstruction methods
interval_width = mat_ancestral_pic$ace - mat_ancestral$ace
node_values = c(rep(NA, nrow(portulacaData)), interval_width)

portulaca_ggtree + 
  geom_nodepoint(aes(color = node_values), size = 3, alpha = 1) +
  scale_color_gradientn(colors = mycol$x)

```

## Mapping character states using a different method
http://www.phytools.org/Cordoba2017/ex/15/Plotting-methods.html 

```{r}

# read in character data again
portulacaData <- read.csv("data/portulaca_means_mapping.csv", header = TRUE, row.names = 1)

portulacaData <- portulacaData %>% select(MAT)

portulacaData <- setNames(portulacaData[,1], rownames(portulacaData))

obj <- contMap(portulacaTree, portulacaData, plot = FALSE)
obj <- setMap(obj, invert = TRUE)
plot(obj, fsize = c(0.5, 0.75), outline = FALSE, lwd = c(3, 7), 
     leg.txt = "Mean annual temperature (C)")

write.csv(obj$cols, "data/color_palette.csv", row.names = FALSE)

```


