---
title: "Trait Evolution for Portulac MAT"
author: "Nora Heaphy"
date: "11/30/2020"
output: html_document
---

```{r setup, include=FALSE}

library(phytools)
library(ape)
library(geiger)
library(caper)
library(Rmpfr)
library(OUwie)
library(qpcR)

knitr::opts_chunk$set(echo = TRUE)
```


#### Data Wrangling

```{r}

# load in tree and data
portulacaTree <- read.tree("data/Ocampo_Portulaca.parts.treefile")

portulacaData <- read.csv("data/portulaca_means_mapping.csv", header = TRUE, 
                          sep = ",", na.strings = "") 
rownames(portulacaData) <- portulacaData$species

portulacaTree$tip.label = portulacaTree$tip.label %>%
  str_replace("-.*","") %>%
  str_replace("_", " ")

# prune tree to include only species we have climate data for
species_character <- row.names(portulacaData) 
portulacaTree <- drop.tip(portulacaTree, setdiff(portulacaTree$tip.label, species_character))
portulacaTree <- drop.tip(portulacaTree, c(9, 4)) # monophyletic duplicates

# put climate data in same order as tip labels
portulacaData <- portulacaData[portulacaTree$tip.label,]

# isolate MAT
port_mat <- portulacaData[,22]
names(port_mat) <- row.names(portulacaData)

# isolate MAP
port_map <- portulacaData[,13]
names(port_map) <- row.names(portulacaData)

```

```{r fig.width=8, fig.height=10}

ggtree(portulacaTree) + 
  geom_tiplab(fontface = "italic") +
  geom_text2(aes(label = node), col = "red", size = 3)

```

```{r}

# plot tip labels for MAT
obj <- contMap(portulacaTree, port_mat, plot = FALSE)
obj <- setMap(obj, invert = TRUE)
plot(obj, lwd = 7)

# plot tip labels for MAP
obj <- contMap(portulacaTree, port_map, plot = FALSE)
obj <- setMap(obj, invert = TRUE)
plot(obj, lwd = 7)

```


#### Estimate Phylogenetic Signal

Phylogenetic signal describes the tendency for related species to resemble each other more than species drawn at random from the same tree. The amount of "phylogenetic signal" in the trait data essentially tells us how much of this observed trait variation is simply due to time since divergence 

The method below uses Blomberg's K to estimate phylogenetic signal. Values of K range from 0 to infinity, with an expected value of 1.0 under Brownian motion. 

Values of K<1.0 describe data with less phylogenetic signal than expected, and values of K>1.0 describe data with greater phylogenetic signal than expected due to chance.

There are other statistics for measuring phylogenetic signal, like Pagel's lambda.

```{r}

phylosig(portulacaTree, port_mat, method = "K")

phylosig(portulacaTree, port_map, method = "K")

```

