---
title: "Ancestral Reconstruction"
author: "Nora Heaphy"
date: "11/15/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ape)
library(ggtree)
library(phytools)
library(tidyverse)
library(ggplot2)
library(here)
library(phylotools)

```

## Data cleaning

- First to get all the taxa names in the character data matrix and the phylogeny tip labels into the format *Genus_species*

- Then drop the tips from the tree that don't have an equivalent in the character data (which is many of them) with `drop.tip`

- Then remove the character data rows that don't have an equivalent in the tree with `dplyr`

- Then use this line from exercise 10 to get the character data rows into the right order to map onto the tips: `geoData = geoData[ match( row.names(geoData), geoTree$tip.label  ), ]`

```{r}

# read in tree file
portulacaTree <- read.tree(here("data", "Ocampo_Portulaca.parts.treefile"))

# read in character data
portulacaData <- read.csv(here("data", "portulaca_means_11-06-2020.csv"), header = TRUE, 
                          sep = ",", na.strings = "")

# remove taxonomy info from tip labels
tip_lab <- data.frame(portulacaTree$tip.label) %>%
  mutate(tip_lab_new = gsub("-.*","", portulacaTree.tip.label)) %>%
  mutate(tip_lab_final = gsub("_", " ", tip_lab_new)) %>%
  select(portulacaTree.tip.label, tip_lab_final)

portulacaTree <- sub.taxa.label(portulacaTree, tip_lab)

# make dataframe of species names from character data
species_character <- data.frame(portulacaData$species) %>%
  rename(names = portulacaData.species)

# make dataframe of species names from tip labels
species_tree <- data.frame(portulacaTree$tip.label) %>%
  rename(names = portulacaTree.tip.label)

# pull out the names in the tree but not the character data
names_to_drop <- setdiff(species_tree, species_character)

```

```{r fig.width=9, fig.height=18}

# plot tree

plot(portulacaTree)

```

